% main=pgfplots.tex

\chapter{Step-by-Step Tutorials}
{%
\tikzset{external/figure name/.add={}{tutorials_}}%


\section{Introduction}

Visualization of data is often necessary and convenient in order to analyze and
communicate results of research, theses, or perhaps just results.

\PGFPlots{} is a visualization tool. The motivation for \PGFPlots{} is that you
as end-user provide the data and the descriptions as input, and \PGFPlots{}
takes care of rest such as choosing suitable scaling factors, scaling to a
prescribed target dimension, choosing a good displayed range, assigning tick
positions, drawing an axis with descriptions placed at appropriate places.

\PGFPlots{} is a solution for an old problem of visualization in \LaTeX{}: its
descriptions use the same fonts as the embedding text, with exactly the same
font sizes. Its direct embedding in \LaTeX{} makes the use of \LaTeX{}'s
powerful math mode as easy as possible: for any kind of axis descriptions up to
user-defined annotations. It features document-wide line-styles, color schemes,
markers\ldots{} all that makes up consistency.

\PGFPlots{} offers high-quality. At the same time, it is an embedded solution:
it is largely independent of 3rd party tools, although it features import
functions to benefit from available tools.

Its main goal is: you provide your data and your descriptions -- and
\PGFPlots{} runs without more input. If you want, you can customize what you
want.


\input{pgfplots.tutorial1.tex}
\input{pgfplots.tutorial2.tex}
\input{pgfplots.tutorial3.tex}
\input{pgfplots.tutorial4.tex}
}%
\endinput   \todosp{does this "endinput" have a special meaning?}
